---
title: Data analysis
subtitle: Age, vocal learning and FoxP2 in budgerigards
author: Marcelo Araya-Salas
date: "`r Sys.Date()`"
toc: true
toc-depth: 2
toc-location: left
number-sections: true
highlight-style: pygments
format:
  html:
    df-print: kable
    code-fold: true
    code-tools: true
    css: qmd.css
editor_options: 
  chunk_output_type: console
---

<!-- skyblue box -->
<div class="alert alert-info">

Data analysis for the paper:

B Moussaoui, K Ulmer, M Araya-Salas, TF Wright. In review. **Persistent vocal learning in an aging open-ended learner reflected in neural FoxP2 expression**


</div>

&nbsp; 

<!-- this code add line numbers to code blocks -->
<!-- only works when code folding is not used in yaml (code_folding: show) -->

```{=html}
<style>
body
  { counter-reset: source-line 0; }
pre.numberSource code
  { counter-reset: none; }
</style>
```

```{r set root directory, echo = FALSE}

knitr::opts_knit$set(root.dir = "..")

```

```{r add link to github repo, echo = FALSE, results='asis'}

# print link to github repo if any
if (file.exists("./.git/config")){
  config <- readLines("./.git/config")
  url <- grep("url",  config, value = TRUE)
  url <- gsub("\\turl = |.git$", "", url)
  cat("\nSource code and data found at [", url, "](", url, ")", sep = "")
  }

```


```{r setup style, echo = FALSE, message = FALSE, warning=FALSE}

# options to customize chunk outputs
knitr::opts_chunk$set(
  class.source = "numberLines lineAnchors", # for code line numbers
  tidy.opts = list(width.cutoff = 65), 
  tidy = TRUE,
  message = FALSE,
  warning = FALSE
 )

```


<!-- skyblue box -->
<div class="alert alert-info">

# Purpose

- Testing for correlation between vocal measures and FoxP2 MMST/VSP Protein Expression and Expression Difference with Age

</div>

&nbsp;

# Load packages

```{r load packages}

# knitr is require for creating html/pdf/word reports
# load function from sketchy
source("https://raw.githubusercontent.com/maRce10/sketchy/main/R/load_packages.R")

# install/ load packages
load_packages(packages = c("ggplot2", "tidyverse", "cowplot", "pwr", "brms", "viridis"))

my.viridis <- function(...) viridis(alpha = 0.5, begin = 0.3, end = 0.7,
    ...)

source("https://raw.githubusercontent.com/maRce10/brmsish/master/R/extended_summary.R")
source("https://raw.githubusercontent.com/maRce10/brmsish/master/R/helpers.R")
source("https://raw.githubusercontent.com/maRce10/brmsish/master/R/check_rds_fits.R")

```


# Load data

```{r}
voc_brain_corr <- read.csv("./data/raw/Vocal_Brain_Corr_03_09_2023_Final.csv", header = TRUE)
head(voc_brain_corr)

```

We can take 1 - acoustic.overlap, acoustic.overlap being calculated as the intersect over union of an individual’s beginning and ending acoustic spaces, to be able to represent vocal plasticity such that higher values representing a more plastic vocal repertoire.
    
    - vocal.plasticity = (1-acoustic.overlap)

```{r}

voc_brain_corr$vocal.plasticity <- (1 - (voc_brain_corr$acoustic.overlap))

```

# Exploratory graphs

```{r}

voc_brain_corr$treatment <-
    factor(voc_brain_corr$treatment, levels = c("Young", "Old"))

#Visualize data
theme_set(theme_classic())
area.change <- ggplot(voc_brain_corr,
                      aes(x = mmst.vsp,
                          y = acoustic.area, colour = treatment)) +
    geom_point(size = 6) + geom_smooth(method = "lm", se = FALSE, size = 1.5) + labs(y = "Vocal diversity", x = "FoxP2 MMSt/VSP expression", color = "Adult age") +  theme(
        text = element_text(size = 20),
        axis.line = element_line(colour = 'black', size = 1.5),
        axis.ticks = element_line(colour = "black", size = 1.5),
        axis.text = element_text(colour = "black")
    ) + scale_color_viridis_d(begin = 0.2, end = 0.8, alpha = 0.5)  + coord_cartesian(xlim = c(0.25, 1.25))
area.change

ggsave(
    "./output/FOXP2_VocalDiversity_Correlation_Final_03_09_2023.png",
    width = 7,
    height = 6,
    units = "in"
)

overlap.self <-
    ggplot(voc_brain_corr,
           aes(x = mmst.vsp, y = vocal.plasticity, colour = treatment)) +
    geom_point(size = 6) + geom_smooth(method = "lm", se = FALSE, size = 1.5) + labs(y = "Vocal plasticity", x = "FoxP2 MMSt/VSP expression", color = "Adult age") +  theme(
        text = element_text(size = 20),
        axis.line = element_line(colour = 'black', size = 1.5),
        axis.ticks = element_line(colour = "black", size = 1.5),
        axis.text = element_text(colour = "black")
    ) + scale_color_viridis_d(begin = 0.2, end = 0.8, alpha = 0.5)  + coord_cartesian(xlim = c(0.25, 1.25), ylim = c(0, 1))
overlap.self 

ggsave(
    "./output/FOXP2_VocalPlasticity_Correlation_Final__03_09_2023.png",
    width = 7,
    height = 6,
    units = "in"
)

overlap.group <- ggplot(voc_brain_corr,
                        aes(x = mmst.vsp,
                            y = acoustic.overlap.to.group,
                            colour = treatment)) +
    geom_point(size = 6) + geom_smooth(method = "lm", se = FALSE, size = 1.5) + labs(y = "Vocal convergence", x = "FoxP2 MMSt/VSP expression", color = "Adult age") +  theme(
        text = element_text(size = 20),
        axis.line = element_line(colour = 'black', size = 1.5),
        axis.ticks = element_line(colour = "black", size = 1.5),
        axis.text = element_text(colour = "black")
    ) + scale_color_viridis_d(begin = 0.2, end = 0.8, alpha = 0.5)  + coord_cartesian(xlim = c(0.25, 1.25), ylim = c(0, 1))
overlap.group

ggsave(
    "./output/FOXP2_VocalConvergence_Correlation_Final__03_03_2023.png",
    width = 7,
    height = 6,
    units = "in"
)

```


# Statistical analysis

## Models without age as a predictor

### Fit models 
```{r, eval = FALSE}

iter <- 20000
chains <- 4
priors <- c(prior(normal(0, 6), class = "b"))

# mean center expression
voc_brain_corr$mmst.vsp_cntr <- voc_brain_corr$mmst.vsp - mean(voc_brain_corr$mmst.vsp)


# cmdstanr::set_cmdstan_path('~/Documentos/cmdstan/')

# to run within-chain parallelization
model.01 <- brm(formula = acoustic.area ~ mmst.vsp_cntr, data = voc_brain_corr,
    prior = priors, iter = iter, chains = chains, cores = chains,
    control = list(adapt_delta = 0.99, max_treedepth = 15), file = "./data/processed/regression_model_acoustic_area_by_expression.RDS",
    file_refit = "always")

model.02 <- brm(formula = vocal.plasticity ~ mmst.vsp_cntr, data = voc_brain_corr,
    prior = priors, iter = iter, chains = chains, cores = chains, family = Beta(),
    control = list(adapt_delta = 0.99, max_treedepth = 15), file = "./data/processed/regression_model_vocal_plasticity_by_expression.RDS",
    file_refit = "always")

model.03 <- brm(formula = acoustic.overlap.to.group ~ mmst.vsp_cntr, data = voc_brain_corr,
    prior = priors, iter = iter, chains = chains, cores = chains, family = Beta(),
    control = list(adapt_delta = 0.99, max_treedepth = 15), file = "./data/processed/regression_model_vocal_overlap_to_group_by_expression.RDS",
    file_refit = "always")

```

### Change in acoustic area (diversity)
```{r, results='asis', fig.height=3}
mod <- readRDS("./data/processed/regression_model_acoustic_area_by_expression.RDS")

extended_summary(mod, n.posterior = 1000, fill = viridis(10)[7], trace.palette = my.viridis,
    remove.intercepts = TRUE, highlight = TRUE, print.name = FALSE)

```


### Overlap of acoustic area to self (plasticity) 
```{r, results='asis', fig.height=3}

mod <- readRDS("./data/processed/regression_model_vocal_plasticity_by_expression.RDS")

extended_summary(mod, n.posterior = 1000, fill = viridis(10)[7], trace.palette = my.viridis,
    remove.intercepts = TRUE, highlight = TRUE, print.name = FALSE)

```

### Overlap to group acoustic area (convergence) 
```{r, results='asis', fig.height=3}

mod <- readRDS("./data/processed/regression_model_vocal_overlap_to_group_by_expression.RDS")

extended_summary(mod, n.posterior = 1000, fill = viridis(10)[7], trace.palette = my.viridis,
    remove.intercepts = TRUE, highlight = TRUE, print.name = FALSE)

```


## Models with foxp2 expression & age as predictors

### Fit models
```{r, eval = FALSE}

# to run within-chain parallelization
model.01 <- brm(formula = acoustic.area ~ mmst.vsp_cntr * treatment, data = voc_brain_corr,
    prior = priors, iter = iter, chains = chains, cores = chains,
    control = list(adapt_delta = 0.99, max_treedepth = 15), file = "./data/processed/regression_model_acoustic_area_by_expression_and_treatment.RDS",
    file_refit = "always")

model.02 <- brm(formula = vocal.plasticity ~ mmst.vsp_cntr * treatment, data = voc_brain_corr,
    prior = priors, iter = iter, chains = chains, cores = chains, family = Beta(),
    control = list(adapt_delta = 0.99, max_treedepth = 15), file = "./data/processed/regression_model_vocal_plasticity_by_expression_and_treatment.RDS",
    file_refit = "always")

model.03 <- brm(formula = acoustic.overlap.to.group ~ mmst.vsp_cntr * treatment, data = voc_brain_corr,
    prior = priors, iter = iter, chains = chains, cores = chains, family = Beta(),
    control = list(adapt_delta = 0.99, max_treedepth = 15), file = "./data/processed/regression_model_vocal_overlap_to_group_by_expression_and_treatment.RDS",
    file_refit = "always")


```

### Change in acoustic area (diversity)
```{r, results='asis', fig.height=3}
mod <- readRDS("./data/processed/regression_model_acoustic_area_by_expression_and_treatment.RDS")

extended_summary(mod, n.posterior = 1000, fill = viridis(10)[7], trace.palette = my.viridis,
    remove.intercepts = TRUE, highlight = TRUE, print.name = FALSE)

```


### Overlap of acoustic area to self (plasticity) 
```{r, results='asis', fig.height=3}

mod <- readRDS("./data/processed/regression_model_vocal_plasticity_by_expression_and_treatment.RDS")

extended_summary(mod, n.posterior = 1000, fill = viridis(10)[7], trace.palette = my.viridis,
    remove.intercepts = TRUE, highlight = TRUE, print.name = FALSE)

```

### Overlap to group acoustic area (convergence) 
```{r, results='asis', fig.height=3}

mod <- readRDS("./data/processed/regression_model_vocal_overlap_to_group_by_expression_and_treatment.RDS")

extended_summary(mod, n.posterior = 1000, fill = viridis(10)[7], trace.palette = my.viridis,
    remove.intercepts = TRUE, highlight = TRUE, print.name = FALSE)

```

## Combined model metadata

```{r, results='asis'}
check_rds_fits(path = "./data/processed/", html = TRUE)
```

<div class="alert alert-success">

# Takeaways {.unnumbered .unlisted}

- Lower diversity in old birds
</div>

&nbsp;

<!-- '---' adds a gray vertical line -->

---

## Posterior predictive checks

Check reilability of Bayesian models 
```{r, results='asis'}
model_list <- c(divergence = "./data/processed/regression_model_acoustic_area_by_expression.RDS",
    plasticity = "./data/processed/regression_model_vocal_plasticity_by_expression.RDS",
    convergence = "./data/processed/regression_model_vocal_overlap_to_group_by_expression.RDS",
    divergence_interaction = "./data/processed/regression_model_acoustic_area_by_expression_and_treatment.RDS",
    plasticity_interaction = "./data/processed/regression_model_vocal_plasticity_by_expression_and_treatment.RDS",
    convergence_interaction = "./data/processed/regression_model_vocal_overlap_to_group_by_expression_and_treatment.RDS")

ndraws <- 20

for (i in seq_len(length(model_list))) {
    cat("\n\n## ", names(model_list)[i], "\n\n")

    fit <- readRDS(model_list[[i]])
    ppc_dens <- pp_check(fit, ndraws = ndraws, type = "dens_overlay")  # shows dens_overlay plot by default
    pp_mean <- pp_check(fit, type = "stat", stat = "mean", ndraws = ndraws)
    pp_scat <- pp_check(fit, type = "scatter_avg", ndraws = ndraws)
    pp_stat2 <- pp_check(fit, type = "stat_2d", ndraws = ndraws)
    pp_loo <- pp_check(fit, type = "loo_pit_qq", ndraws = ndraws)
    pp_error <- pp_check(fit, type = "error_scatter_avg", ndraws = ndraws)
    plot_l <- list(ppc_dens, pp_mean, pp_scat, pp_error, pp_stat2,
        pp_loo)

    plot_l <- lapply(plot_l, function(x) x + scale_color_viridis_d(begin = 0.1,
        end = 0.8, alpha = 0.5) + scale_fill_viridis_d(begin = 0.1,
        end = 0.8, alpha = 0.5) + theme_classic())

    print(plot_grid(plotlist = plot_l, ncol = 2))
}
```


&nbsp; 
 
 <!-- add packages used, system details and versions  -->
 
<font size="4">Session information</font>

```{r session info, echo=F}

sessionInfo()

```
